A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-15 latex/main.tex
Move the USHER+RIPPLES section to the tend
Also comment out section on compute

Update

\subsection*{Comparison with UShER+RIPPLES}

RIPPLES~\cite{Turakhia2022} is a method for identifying
recombination events using a phylogeny inferred by UShER.
This distinguishes RIPPLES from other recombination detection
methods such as 3SEQ, which classify particular \emph{sequences}
as recombinant or not and can only identify potential onward 
transmission post-hoc. The combination of UShER and RIPPLES
is therefore the only method that is directly comparable 
to sc2ts. It works by first identifying plausible long 
branches in the input tree.
For each candidate branch,
RIPPLES systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.
A recombination event at that node is then flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny
(the ``parsimony improvement'' parameter, $p=3$, by default).
RIVET~\cite{Smith2023tracking} is a tool for quality control and
visualization of these recombination events.

To compare the accuracy of sc2ts with UShER+RIPPLES
we ran RIPPLES the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
We used the RIPPLES parameter $p=4$ as is equivalent to the
sc2ts $k=4$ parameter used to build the ARG,
which resulted in 11[XX] distinct recombination events.
We then ran 3SEQ on each of the candidate trios,
giving an overall validation rate of XX\% (STAR Methods).
Comparing events identified in the UShER tree and sc2ts ARG
is not straightforward in general, and
so we focus comparison on the singleton recombinants
returned by both.
Although these are likely to be enriched for artefacts,
both methods face the same challenges and
the events can be compared like-for like.
Of the 11XX recombination events returned by RIPPLES,
757 are singletons (respectively, 525/855 for sc2ts).
Of the RIPPLES singletons, 705 are also singletons in the sc2ts ARG.
[There's a complicated story to tell here
about extra singletons in the RIPPLES set which pass
3SEQ validation, but have parents from the future.]
Thus, 216 of these recombinants pass 3SEQ validation,
have at least one parent that is on average
63 days \emph{younger} than the sequence itself.
These parents were not available to sc2ts as potential matches,
as it enforces strict time ordering based on sample metadata.
Further study is required to determine
if the metadata is incorrect in these cases, or
if these should be flagged as false positives.
However, the key point is that RIPPLES is dependent on signals
embedded in the UShER tree, and therefore we cannot
decouple the finer details of post-hoc recombinant detection
from the tree building algorithm.

We also compared the recombination events
output by RIPPLES with the Pango X lineages.
At p=4, we found that only XX of the Pango X lineages were present and
many well-characterized events such as XBB were not present.
We therefore reran at p=3, resulting in XXXX events overall.
At p=4, events involving XX of the 44 Pango X lineages
present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages other lineages associated with two competing events.
[More summary]

[Paragraph summing things up]
See Document~\ref{sec:detailed_ripples} for more detailed analysis 
of the UShER+RIPPLES results.

@@

\subsection*{Efficient inference and analysis of a pandemic-scale ARG}

The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
which overwhelmed existing computational approaches,
motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
the functionality required largely overlaps.
Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
and the VCF Zarr data storage format \cite{Czech2025}
to describe relationships among virus samples,
building an ARG over time as data becomes available.
By reusing high-quality and high-performance components of these toolkits,
we could develop the sc2ts inference method itself and
perform all the pandemic-scale analyses in this article
with Python code alone and no compiled languages.

To illustrate the power of these libraries, we provide some examples
in which we compare the Usher and sc2ts intersections in tskit format described earlier.
First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
The total aligned dataset for 1,226,587 samples and 27,431 sites
shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
Of the 80,574,518 missing data calls (Ns) in the alignments,
sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
Of these sc2ts imputed only 77 (0.02\%) incorrectly
(i.e., with a base that is not compatible with the ambiguity code).
UShER imputed only 4 calls from this set incorrectly,
reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
This computation,
performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
with a peak memory usage of about 4.5 GB
(see \cite{Czech2025} for further benchmarks for
processing the Viridian data in VCF Zarr format).
Performing this analysis using matUtils would require
exporting the full dataset to VCF, and
converting the VCF to FASTA, and
then comparing the alignments one-by-one to the source FASTA alignments.
In contrast with matUtils and most phylogenetic software,
which provide a command line interface and suite of predefined operations that typically output text files,
tskit is a Python API (with C and Rust interfaces also supported)
which has many advantages \cite{Baumdicker2022}.
As an illustration of the open-ended and efficient nature of the analyses this enables,
we computed the number of polytomies (nodes that have $> 2$ children)
and the maximum number of children per node in the Usher and sc2ts outputs.
In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
and compute the maximum number of children (5918).
In the sc2ts ARG, it took 385 ms and 6 lines of code
to find the maximum number of polytomies (137,001) and
maximum number of children (7708) across all 348 local trees.

>>>>

@@

\subsection*{Comparison with UShER+RIPPLES}
RIPPLES~\cite{Turakhia2022} is a method for identifying
recombination events using a phylogeny inferred by UShER.
This distinguishes RIPPLES from other recombination detection
methods such as 3SEQ, which classify particular \emph{sequences}
as recombinant or not and can only identify potential onward transmission post-hoc. 
The combination of UShER and RIPPLES
is therefore the only method that is directly comparable to sc2ts. 
RIPPLES works by first identifying long 
branches in the input tree, by default those carrying 3 or more 
mutations. For each candidate branch,
RIPPLES systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.
A recombination event at that node is then flagged if
the overall parsimony of these multiple placements
sufficiently improves over
the original single-parent origin in the phylogeny.
This ``parsimony improvement'' parameter ($p$)
is equivalent to the sc2ts recombination penalty ($k$).

To compare the performance of sc2ts and UShER+RIPPLES 
in terms of recombinant detection, we ran RIPPLES the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
Using the RIPPLES parameter $p=4$ (equivalent to $k=4$ used 
to build the sc2ts ARG) resulted in 1174 reported recombination events.
We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
were classified as recombinant, giving an overall validation rate of 51\%.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both, 
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination in the sc2ts ARG can largely
be explained by a combination of two factors.
Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. Thus, the parental
sequences were simply not available to match against for 
sc2ts (which enforces strict time ordering) when the 
recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occured at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER
and sc2ts.

The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterised events such as XBB.
We also compared the recombination events
output by RIPPLES with the Pango X lineages.
At p=4, we found that only XX of the Pango X lineages were present and
many well-characterized events such as XBB were not present.
We therefore reran at p=3, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%).
At p=3, events involving XX of the 44 Pango X lineages
present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages other lineages associated with two competing events.
[More summary]

[Paragraph summing things up. It seems like RIPPLES is doing its
job, and the question is really whether UShER is giving 
the correct signals for it to detect. Sc2ts detects more of the 
well-known events. Needs careful explanation.].
See Document~\ref{sec:detailed_ripples} for more detailed analysis 
of the UShER+RIPPLES results.

% TODO convert this into a discussion pargraph..
% \subsection*{Efficient inference and analysis of a pandemic-scale ARG}

% The COVID-19 pandemic resulted in the collection of viral genome sequence data on an unprecedented scale,
% which overwhelmed existing computational approaches,
% motivating the rapid development of new methods \cite{Turakhia2021,DeMaio2023,Hinrichs2024}.
% Although the computational tools used in phylogenetics and population genetics have historically been largely distinct,
% the functionality required largely overlaps.
% Sc2ts was built by modifying components of the tskit ARG software ecosystem \cite{Kelleher2019,Ralph2020,Baumdicker2022,Wong2024}
% and the VCF Zarr data storage format \cite{Czech2025}
% to describe relationships among virus samples,
% building an ARG over time as data becomes available.
% By reusing high-quality and high-performance components of these toolkits,
% we could develop the sc2ts inference method itself and
% perform all the pandemic-scale analyses in this article
% with Python code alone and no compiled languages.

% To illustrate the power of these libraries, we provide some examples
% in which we compare the Usher and sc2ts intersections in tskit format described earlier.
% First, we computed levels of concordance when imputing missing and ambiguous bases in the alignments.
% The total aligned dataset for 1,226,587 samples and 27,431 sites
% shared by the sc2ts and Usher tree sequences represents 31.34 GB of nucleotide calls.
% Of the 80,574,518 missing data calls (Ns) in the alignments,
% sc2ts and Usher disagreed in their imputed values for only 56,782 sites (0.07\%).
% Additionally, 413,303 calls made use of the IUPAC uncertainty codes.
% Of these sc2ts imputed only 77 (0.02\%) incorrectly
% (i.e., with a base that is not compatible with the ambiguity code).
% UShER imputed only 4 calls from this set incorrectly,
% reflecting its explicit incorporation of IUPAC uncertainty codes in parsimony calculations.
% This computation,
% performed by co-iterating over the sites in the alignments and the sc2ts and Usher trees,
% required 30 lines of Python code and took 2m 59s on an intel i7-9700 CPU,
% with a peak memory usage of about 4.5 GB
% (see \cite{Czech2025} for further benchmarks for
% processing the Viridian data in VCF Zarr format).
% Performing this analysis using matUtils would require
% exporting the full dataset to VCF, and
% converting the VCF to FASTA, and
% then comparing the alignments one-by-one to the source FASTA alignments.
% In contrast with matUtils and most phylogenetic software,
% which provide a command line interface and suite of predefined operations that typically output text files,
% tskit is a Python API (with C and Rust interfaces also supported)
% which has many advantages \cite{Baumdicker2022}.
% As an illustration of the open-ended and efficient nature of the analyses this enables,
% we computed the number of polytomies (nodes that have $> 2$ children)
% and the maximum number of children per node in the Usher and sc2ts outputs.
% In the Usher tree it took 116 ms and 4 lines of code to count polytomies (171,148)
% and compute the maximum number of children (5918).
% In the sc2ts ARG, it took 385 ms and 6 lines of code
% to find the maximum number of polytomies (137,001) and
% maximum number of children (7708) across all 348 local trees.
